Algorithms for the Automated Correction of Vertical Drift in Eye Tracking Data

### 2020-10-23 manuscript/manuscript.tex
Strip out highlighting of new material

% HIGHLIGHTING OF NEW MATERIAL
\usepackage{xcolor}
\newcommand\NEWMATERIAL[1]{\textcolor{blue}{#1}}
% HIGHLIGHTING OF NEW MATERIAL

\title{Algorithms for the Automated Correction of Vertical Drift in Eye Tracking Data}
\shorttitle{Algorithms for the correction of vertical drift}

@@

A technical issue that arises from the particular circumstances of multiline reading is so-called ``vertical drift,'' which we define as the progressive displacement of fixation registrations on the vertical axis \textit{over time}. In other words, fixations may be recorded above or below the line of text that the participant was actually reading, \NEWMATERIAL{and the degree and directionality of this error may fluctuate dynamically with each subsequent fixation}, making it nontrivial to eliminate. Fig.~\ref{fig01}A depicts a reading trial exhibiting vertical drift phenomena; in this case, fixations---especially those on the left hand side---are recorded around one line higher than where the reader was actually fixating, but they also tend to slope down to the right such that fixations on the right hand side seem to be better aligned.

Vertical drift can occur quite unpredictably, even following good quality calibration, \NEWMATERIAL{and it is likely caused by spatial phenomena such as degraded eye tracker calibration at the extremes of the screen, or temporal phenomena such as subtle movements in head position or pupil dilation, which can be difficult to control for, even in a laboratory setting. Such sources of measurement error are often exacerbated in the context of multiline reading because, in comparison to single words or sentences, passages of text are distributed over a larger portion of the screen, including areas where general calibration may be worse, and they take longer to read, during which time calibration may begin to degrade}. There are also less frequent opportunities to recalibrate the device during passage reading, since this can only be performed between trials or pages and not during the reading of a passage.

@@

To be clear, the algorithms we consider in this paper are restricted to one specific problem. \NEWMATERIAL{Firstly, we only consider algorithms designed for the ordinary reading of passages of text; other uses of eye tracking, such as visual search and scene perception, can also undergo drift correction, but the methods required are quite different \parencite[see e.g.,][]{Zhang:2011,Zhang:2014}. Similarly, the reading of source code has received some attention, but the affordances and constraints in this domain are quite different from ordinary linguistic reading \parencite[see e.g.,][]{Palmer:2016,Nussli:2011}.} Secondly, we only consider the problem of post-hoc correction; vertical drift can also be corrected in real time, but this imposes a more restrictive set of constraints that are better handled by other types of algorithm \parencite[see e.g.,][]{Hyrskykari:2006,Sibert:2000}. Thirdly, we only consider fully automated algorithms that do not require human supervision.

@@

The \texttt{compare} algorithm is directly based on the method reported by \textcite{LimaSanches:2015} \NEWMATERIAL{and is very similar to the more complex methods described by \textcite{Yamaya:2017}}. The fixation sequence is first segmented into ``gaze lines'' by identifying the return sweeps---long saccades that move the eye from the end of one line to the start of the next. The algorithm considers any saccade that moves from right to left by more than some threshold value (we use 512~px) to be a return sweep. Gaze lines are then matched to text lines based on a measure of similarity between them. \textcite{LimaSanches:2015} considered three measures of similarity and found dynamic time warping \parencite[DTW;][]{Vintsyuk:1968,Sakoe:1978} to be the best method \NEWMATERIAL{(we discuss DTW in more detail later in this section)}. \NEWMATERIAL{Similarly, \textcite{Yamaya:2017} use the closely related Needleman--Wunsch algorithm \parencite{Needleman:1970}.}

The gaze lines and text lines are compared in terms of their \textit{x}~values under the assumption that the fixations in a gaze line should have a good horizontal alignment with the centers of the words in the corresponding text line. Relying only on the \textit{x}~values helps the algorithm overcome vertical drift issues, but it is also problematic because in many standard reading scenarios the lines of text in a passage tend to be horizontally similar to each other; each line tends to contain a similar number of words that are of a similar length, resulting in potential ambiguity about how gaze lines and text lines should be matched up. To alleviate this issue, both \textcite{LimaSanches:2015} \NEWMATERIAL{and \textcite{Yamaya:2017}} only compare the gaze line to a certain number of nearby text lines (we set this parameter to 3, \NEWMATERIAL{which is effectively the closest line plus one line above and one line below}).

@@

\NEWMATERIAL{\texttt{regress} tracks FixAlign very closely, except that we did not implement the ``run rule,'' an option that is switched on by default in FixAlign. This option maps ambiguous fixations to the same line as the surrounding fixations, if the surrounding fixations were classified unambiguously \parencite[p.~680]{Cohen:2013}. Cohen's run rule is a more general method that could in principle be applied to the output of any algorithm, so in the interest of isolating the core concept of FixAlign and comparing all algorithms on a level playing field, we did not to implement the option here.}

@@

As far as we know, the \texttt{split} algorithm takes an approach that is distinct from anything previously reported, although it is conceptually similar to \texttt{segment}. Like \texttt{segment}, the \texttt{split} algorithm works on the principle of splitting the fixation sequence into subsequences by first identifying the return sweeps. However, \texttt{split} is not restricted to finding exactly $m-1$ return sweeps; instead, it identifies the most likely set of return sweeps, however many that turns out to be. There are various ways of approaching this classification problem, but here we use \textit{k}-means clustering to partition the set of saccades into exactly two clusters. Since return sweeps are usually highly divergent from normal saccades \NEWMATERIAL{(i.e., a return sweep is usually represented by a large negative change on the \textit{x}-axis)}, one of the two clusters will invariably contain the return sweeps, which can then be used to split the fixation sequence into subsequences. However, since this is not guaranteed to produce $m-1$ return sweeps (and therefore $m$ subsequences), an order-based mapping is not possible, so \texttt{split} must use absolute position: Subsequences are mapped to the closest text lines in absolute terms. \texttt{split} has the advantage of generally finding all true return sweeps, and even if it identifies some false positives, the resulting subsequences can still be mapped to the appropriate lines by absolute position. However, this also means the algorithm is less resilient to vertical drift issues.

@@

\NEWMATERIAL{The \texttt{stretch} algorithm is loosely based on the the method proposed by \textcite{Lohmeier:2015} and shares some similarities with \textcite{MartinezGomez:2012} and \textcite{Nussli:2011}. \possessivecite{Lohmeier:2015} original method was designed for the reading of source code and therefore takes advantage of the fact that code has very irregular line lengths and indentation levels. The method works by finding an \textit{x}-offset, \textit{y}-offset, and scaling factor that, once applied to the fixations, minimizes alignment error between the fixations and lines of text.}

\NEWMATERIAL{The framework we adopt herein never adjusts the \textit{x}~values, and we also assume that an ordinary passage of text is being read, so line length is substantially more constant than during code reading and therefore less informative. Therefore, we simplified the original method by dispensing with all dependencies on the \textit{x}~values. Instead, \texttt{stretch} finds a \textit{y}-offset, $o^\ast$, and a vertical scaling factor, $s^\ast$, that minimizes the sum absolute difference between the corrected fixation positions ($f_y s + o$) and the corrected fixation positions once attached to their closest lines. The equations presented in \textcite[pp.~37--38]{Lohmeier:2015} therefore simplify to:}

@@

\noindent
\NEWMATERIAL{where $\mathrm{attach}(\cdot)$ returns the \textit{y}-axis position of the nearest line of text. In other words, the algorithm seeks a transformation of the fixations that results in minimal change following the application of \texttt{attach}.}

\NEWMATERIAL{To constrain the minimization problem, the user must specify appropriate lower and upper bounds for the offset and scaling factor, resulting in four free parameters. Here we adopt offset bounds of $[-50, 50]$, following the \texttt{regress} algorithm, and scaling factor bounds of $[0.9, 1.1]$. Effectively, this means the algorithm can move the set of fixations up or down by up~to 50~pixels and stretch their positions on the vertical axis by between 90\% and 110\%. While approaching the problem from a different angle, \texttt{stretch} is computationally similar to \texttt{regress}, except that it emphasizes systematic offset issues rather than systematic slope issues.}

@@

\NEWMATERIAL{DTW is typically useful when you have two sequences, not necessarily of the same length, and you want to (a) calculate how similar they are (as is the case in the \texttt{compare} algorithm) or (b) align the two sequences by mapping each element in one sequence to a corresponding element in the other. For example, DTW may be used to calculate the similarity between a signature, which can be expressed as a sequence of \textit{xy}-coordinates over time, and a reference signature \parencite[e.g.,][]{Lei:2005,Riesen:2018}. Importantly, the two sequences do not need to be perfectly matched in terms of overall magnitude or patterns of acceleration and deceleration for a good alignment to be found. In the case of signature verification, for example, it does not matter if the candidate signature has the same size as the reference or that it was drawn at the same speed, what matters is that there is a good match in the overall shape and that the strokes were drawn in the same order. DTW finds many other applications in, for example, genomics \parencite{Aach:2001}, medicine \parencite{Caiani:1998}, and robotics \parencite{Vakanski:2014}.}

@@

If the final fixation on line~$i$ were mapped to the first word on line~$i+1$, this would result in a large increase in the overall cost of the mapping, so line changes act as major clues about the best alignment. The upshot of this is that \texttt{warp} effectively segments the fixation sequence into exactly $m$ subsequences, which are mapped to the lines of text in chronological order. In this sense, \texttt{warp} behaves very much like \texttt{segment}. However, the additional benefit of \texttt{warp} is that it can simultaneously consider different possibilities about which saccades are the return sweeps, selecting only those that result in the best fit to the passage at a global level. Nevertheless, \texttt{warp} is ultimately limited by the veracity of the expected fixation sequence, which encodes one particular way of reading the passage---line by line from start to end. If the reader deviates from this assumption \NEWMATERIAL{(e.g., by rereading or skipping lines)}, \texttt{warp} can fail to correctly assign fixations to lines.

@@

For each phenomenon, we ran 100 simulations for each of 50 gradations in the parameter space, and each of these 5000 simulated reading scenarios was corrected by all ten algorithms. Accuracy is measured as the percentage of fixations that were correctly mapped back to the target line. Before describing the results, there are three important things to note. Firstly, the extreme values we have chosen for each phenomenon are arbitrary, so the algorithms should only be compared within, and not across, phenomena.\footnote{For example, \texttt{regress} appears to have worse performance on shift compared to slope; however, if we had simulated a narrower range of shift values, the results might have led us to the opposite conclusion.} \NEWMATERIAL{Secondly, we have not modeled the interactions between phenomena because it is inherently difficult to explore the effects of five dimensions on accuracy and it is not clear how the dimensions should be weighted a~priori.} Thirdly, for the algorithms that have free parameters (\texttt{chain}, \texttt{compare}, \texttt{merge}, \texttt{regress}, and \texttt{stretch}), we use the default parameter settings defined in the previous section. We have not systematically manipulated the parameter settings because (a) this would result in an explosion in the number of algorithm/parameter combinations that we must consider, (b) manipulating a parameter to deal with one phenomenon can have unexpected consequences for other phenomena,\footnote{For example, if the standard deviation bounds of \texttt{regress} are widened, it may be possible to improve performance on noise, but the algorithm will be less capable of dealing with slope.} and (c), in a sense, these algorithms ought to incur a penalty for not being parameter-free.

@@

In terms of shift (Fig.~\ref{fig04}C), when the parameter is zero, all algorithms perform perfectly, but as it becomes more extreme, five of the algorithms---\texttt{attach}, \texttt{chain}, \texttt{regress}, \texttt{split}, and \texttt{stretch}---drop in accuracy. In fact, \texttt{attach}, \texttt{chain}, and \texttt{split} produce identical results in the case of shift because they are all fundamentally reliant on absolute position. \NEWMATERIAL{Somewhat surprisingly, \texttt{stretch} does not perform especially well on shift. This is because \texttt{stretch} can only handle up~to one full line of shift; any more than this and the bounds have to be relaxed, but this results in an objective function with multiple maxima which is difficult to optimize.} The \texttt{compare} algorithm is mostly resilient to shift, while the remaining four algorithms---\texttt{cluster}, \texttt{merge}, \texttt{segment}, and \texttt{warp}---are invariant.

@@

The cleaned sample dataset was corrected independently by two human correctors (JWC and VNP). To perform the correction, each corrector studied plots of the participants' fixation sequences and recorded, fixation by fixation, which line each one belonged to, guided by fixation position, saccade trajectories, textual cues, and fixation duration, as well as general knowledge of eye tracking and reading behavior. Unlike the algorithms, the human correctors also had the option to discard fixations as they saw fit. \NEWMATERIAL{This is because there were cases where it was clear a fixation should be discarded---for instance, due to spatial misplacement or ultra-short duration---and it would have therefore felt disingenuous to assign these cases to a line anyway.}

@@

As with the simulations, accuracy is measured as the percentage of fixations that the algorithm mapped to the correct line; the ground truth is defined by the gold standard manual correction. In cases where the correctors chose to discard a fixation, the algorithm is automatically wrong, \NEWMATERIAL{which amounts to a constant baseline level of error that all algorithms suffer from equally}. Fig.~\ref{fig07} plots accuracy on the 48 sample trials by algorithm. The most striking result is \texttt{compare} with overall median accuracy of 61.2\%, substantially worse than all other algorithms.\footnote{This concurs with the 60\% accuracy reported by \textcite[p.~1231]{LimaSanches:2015}. \NEWMATERIAL{\textcite[p.~104]{Yamaya:2017}, who describe a slightly more complex method with better sweep detection, report accuracy of 87\%, which is still quite low compared to the other results obtained in this paper.}} This contrasts with our simulations, which indicated that \texttt{compare} should at~least be relatively strong on distortion. The reason for this discrepancy is that the simulated fixation sequences were generated directly from the lines of text with one fixation per word, so the artificial gaze lines that \texttt{compare} identified tended to be very horizontally similar to the artificial text lines. In the natural dataset, however, this is not the case; when the data contains a~lot of natural noise and regressions, gaze lines cannot be reliably matched to text lines based on similarity, even if the set of candidate text lines is narrowed down to the three closest neighbors. Given that \texttt{compare} exhibited such poor performance, we consider it to be an algorithmic deadend and do not discuss it any further.

@@

\NEWMATERIAL{It is also worth considering what proportion of trials are likely to be usable following the application of a particular method. Fig.~\ref{fig10} reports the proportion of trials that obtained accuracy of 90\%, 95\%, and 99\% by algorithm. If you are willing to accept relatively low accuracy at the trial level (e.g., 90\%, Fig~\ref{fig10}A), then \texttt{cluster}, \texttt{merge}, and \texttt{stretch} will provide the best performance---a large proportion of trials will meet this criterion. In comparison, if you have more stringent accuracy requirements at the trial level (e.g., 99\%, Fig~\ref{fig10}C), then \texttt{segment}, \texttt{split}, and \texttt{warp} are likely to provide better performance. Of course, the cost of a more stringent accuracy criterion is that fewer trials will be usable overall, and it is not possible to know which trials have low accuracy in the absence of manual correction data. This highlights the fact that it is currently not possible to confidently achieve a high level of accuracy in a high proportion of trials, so researchers may still need to invest a significant amount of time if a high level of accuracy is demanded.}

@@

\NEWMATERIAL{The analyses presented in this paper clearly indicate that each algorithm performs best on a different set of factors, some of which we have not considered here in detail. For example, if the line spacing is quite tight, the eye tracking data is more likely to be negatively impacted by distortion, making a sequential algorithm a better choice; conversely, if lines are spread far apart, a relative-positional algorithm may be more appropriate.} Overall, different sets of data will require different correction algorithms, and a qualitative inspection of the data will be required to detect the relative severity of general noise, drift issues, and regression phenomena. To help in this process, one option might be to hand-correct a sample of the trials in order to assess which algorithm performs best on those specific cases and then apply this algorithm to the entire dataset. However, there might very well be too much trial-by-trial (or participant-by-participant) variability to use a single algorithm across an entire dataset. In this case, it may be preferable to create subsets of data exhibiting comparable patterns of eye tracking phenomena and deal with those subsets with different algorithms. \NEWMATERIAL{Another idea would be to run several algorithms over the dataset and manually inspect cases where there is disagreement.}

@@

Another important aspect to consider is the presence of free parameters. The \texttt{chain}, \texttt{merge}, \texttt{regress}, and \texttt{stretch} algorithms take additional input parameters that must be set appropriately by the user. In practice, \possessivecite{Spakov:2019} and \possessivecite{Cohen:2013} suggested defaults for the \texttt{merge} and \texttt{regress} algorithms seemed to work well on our test dataset, but in the case of \texttt{chain}, it was somewhat unclear how to set the \textit{x}~and \textit{y}~thresholds appropriately, so experimentation might be required to produce the best results. \NEWMATERIAL{We also found that \texttt{stretch} was very sensitive to its parameter settings and that the upper and lower bounds must be tightly constrained around likely values for it to produce sensible results.} An advantage of all other algorithms is that they are parameter free, making drift correction easier to perform, document, and justify.

@@

We would not wish to claim that the algorithms, as presented here, are the only approaches one may take nor that they are the ultimate form of each core method; all can be improved in one way or another. \NEWMATERIAL{Furthermore, there are likely to be ways of combining the outputs of multiple algorithms to increase confidence in particular solutions.} The goal of this paper, however, was to evaluate the algorithms in their more abstract, idealized forms in order to make general recommendations and to provide a solid foundation for the future development of vertical drift correction software. Nevertheless, here we briefly note some of the most obvious ways in which the algorithms could be improved.

@@

The biggest weakness of \texttt{cluster} was its ability to deal with general noise \NEWMATERIAL{(or, equivalently, tight line spacing)}. One potential way to improve this would be to utilize the \textit{x}~values of the fixations. Unfortunately, it is not simply a case of performing a two-dimensional \textit{k}-means clustering on the \textit{xy}~values because this leads to situations where clusters are identified that span multiple lines because they have similar \textit{x}~values. However, it might still be possible to utilize the \textit{x}-axis information, perhaps by weighting the two axes differently in some way. Cluster analysis is a very broad topic in data science, and there are likely to be many other candidate algorithms, beyond simple \textit{k}-means clustering, that will be worth investigating.

@@

\NEWMATERIAL{Our analyses showed that \texttt{stretch} behaves very similarly to \texttt{regress}. This is because they are essentially two variants on the same basic idea: Detect the magnitude of the underlying slope (\texttt{regress}) or shift (\texttt{stretch}) and then reverse it. However, this does not work so well if the underlying calibration error is fluctuating in time or space. One way to improve the method, then, would be to search a more complex transformation space by including rotations and shears, for example, or by applying separate transformations to each quadrant of the text. Additionally, as Equation~\ref{equation1} makes clear, \texttt{stretch} essentially has the \texttt{attach} algorithm embedded within it, but in principle it should be possible to substitute this with any of the positional algorithms. For example, a \texttt{stretch-chain} algorithm would find a transformation of the fixations that results in minimal change when you apply the \texttt{chain} algorithm.}

>>>>

\title{Algorithms for the Automated Correction of Vertical Drift in Eye Tracking Data}

\shorttitle{Algorithms for the correction of vertical drift}

@@

A technical issue that arises from the particular circumstances of multiline reading is so-called ``vertical drift,'' which we define as the progressive displacement of fixation registrations on the vertical axis \textit{over time}. In other words, fixations may be recorded above or below the line of text that the participant was actually reading, and the degree and directionality of this error may fluctuate dynamically with each subsequent fixation, making it nontrivial to eliminate. Fig.~\ref{fig01}A depicts a reading trial exhibiting vertical drift phenomena; in this case, fixations---especially those on the left hand side---are recorded around one line higher than where the reader was actually fixating, but they also tend to slope down to the right such that fixations on the right hand side seem to be better aligned.

Vertical drift can occur quite unpredictably, even following good quality calibration, and it is likely caused by spatial phenomena such as degraded eye tracker calibration at the extremes of the screen, or temporal phenomena such as subtle movements in head position or pupil dilation, which can be difficult to control for, even in a laboratory setting. Such sources of measurement error are often exacerbated in the context of multiline reading because, in comparison to single words or sentences, passages of text are distributed over a larger portion of the screen, including areas where general calibration may be worse, and they take longer to read, during which time calibration may begin to degrade. There are also less frequent opportunities to recalibrate the device during passage reading, since this can only be performed between trials or pages and not during the reading of a passage.

@@

To be clear, the algorithms we consider in this paper are restricted to one specific problem. Firstly, we only consider algorithms designed for the ordinary reading of passages of text; other uses of eye tracking, such as visual search and scene perception, can also undergo drift correction, but the methods required are quite different \parencite[see e.g.,][]{Zhang:2011,Zhang:2014}. Similarly, the reading of source code has received some attention, but the affordances and constraints in this domain are quite different from ordinary linguistic reading \parencite[see e.g.,][]{Palmer:2016,Nussli:2011}. Secondly, we only consider the problem of post-hoc correction; vertical drift can also be corrected in real time, but this imposes a more restrictive set of constraints that are better handled by other types of algorithm \parencite[see e.g.,][]{Hyrskykari:2006,Sibert:2000}. Thirdly, we only consider fully automated algorithms that do not require human supervision.

@@

The \texttt{compare} algorithm is directly based on the method reported by \textcite{LimaSanches:2015} and is very similar to the more complex methods described by \textcite{Yamaya:2017}. The fixation sequence is first segmented into ``gaze lines'' by identifying the return sweeps---long saccades that move the eye from the end of one line to the start of the next. The algorithm considers any saccade that moves from right to left by more than some threshold value (we use 512~px) to be a return sweep. Gaze lines are then matched to text lines based on a measure of similarity between them. \textcite{LimaSanches:2015} considered three measures of similarity and found dynamic time warping \parencite[DTW;][]{Vintsyuk:1968,Sakoe:1978} to be the best method (we discuss DTW in more detail later in this section). Similarly, \textcite{Yamaya:2017} use the closely related Needleman--Wunsch algorithm \parencite{Needleman:1970}.

The gaze lines and text lines are compared in terms of their \textit{x}~values under the assumption that the fixations in a gaze line should have a good horizontal alignment with the centers of the words in the corresponding text line. Relying only on the \textit{x}~values helps the algorithm overcome vertical drift issues, but it is also problematic because in many standard reading scenarios the lines of text in a passage tend to be horizontally similar to each other; each line tends to contain a similar number of words that are of a similar length, resulting in potential ambiguity about how gaze lines and text lines should be matched up. To alleviate this issue, both \textcite{LimaSanches:2015} and \textcite{Yamaya:2017} only compare the gaze line to a certain number of nearby text lines (we set this parameter to 3, which is effectively the closest line plus one line above and one line below).

@@

\texttt{regress} tracks FixAlign very closely, except that we did not implement the ``run rule,'' an option that is switched on by default in FixAlign. This option maps ambiguous fixations to the same line as the surrounding fixations, if the surrounding fixations were classified unambiguously \parencite[p.~680]{Cohen:2013}. Cohen's run rule is a more general method that could in principle be applied to the output of any algorithm, so in the interest of isolating the core concept of FixAlign and comparing all algorithms on a level playing field, we did not to implement the option here.

@@

As far as we know, the \texttt{split} algorithm takes an approach that is distinct from anything previously reported, although it is conceptually similar to \texttt{segment}. Like \texttt{segment}, the \texttt{split} algorithm works on the principle of splitting the fixation sequence into subsequences by first identifying the return sweeps. However, \texttt{split} is not restricted to finding exactly $m-1$ return sweeps; instead, it identifies the most likely set of return sweeps, however many that turns out to be. There are various ways of approaching this classification problem, but here we use \textit{k}-means clustering to partition the set of saccades into exactly two clusters. Since return sweeps are usually highly divergent from normal saccades (i.e., a return sweep is usually represented by a large negative change on the \textit{x}-axis), one of the two clusters will invariably contain the return sweeps, which can then be used to split the fixation sequence into subsequences. However, since this is not guaranteed to produce $m-1$ return sweeps (and therefore $m$ subsequences), an order-based mapping is not possible, so \texttt{split} must use absolute position: Subsequences are mapped to the closest text lines in absolute terms. \texttt{split} has the advantage of generally finding all true return sweeps, and even if it identifies some false positives, the resulting subsequences can still be mapped to the appropriate lines by absolute position. However, this also means the algorithm is less resilient to vertical drift issues.

@@

The \texttt{stretch} algorithm is loosely based on the the method proposed by \textcite{Lohmeier:2015} and shares some similarities with \textcite{MartinezGomez:2012} and \textcite{Nussli:2011}. \possessivecite{Lohmeier:2015} original method was designed for the reading of source code and therefore takes advantage of the fact that code has very irregular line lengths and indentation levels. The method works by finding an \textit{x}-offset, \textit{y}-offset, and scaling factor that, once applied to the fixations, minimizes alignment error between the fixations and lines of text.

The framework we adopt herein never adjusts the \textit{x}~values, and we also assume that an ordinary passage of text is being read, so line length is substantially more constant than during code reading and therefore less informative. Therefore, we simplified the original method by dispensing with all dependencies on the \textit{x}~values. Instead, \texttt{stretch} finds a \textit{y}-offset, $o^\ast$, and a vertical scaling factor, $s^\ast$, that minimizes the sum absolute difference between the corrected fixation positions ($f_y s + o$) and the corrected fixation positions once attached to their closest lines. The equations presented in \textcite[pp.~37--38]{Lohmeier:2015} therefore simplify to:

@@

\noindent
where $\mathrm{attach}(\cdot)$ returns the \textit{y}-axis position of the nearest line of text. In other words, the algorithm seeks a transformation of the fixations that results in minimal change following the application of \texttt{attach}.

To constrain the minimization problem, the user must specify appropriate lower and upper bounds for the offset and scaling factor, resulting in four free parameters. Here we adopt offset bounds of $[-50, 50]$, following the \texttt{regress} algorithm, and scaling factor bounds of $[0.9, 1.1]$. Effectively, this means the algorithm can move the set of fixations up or down by up~to 50~pixels and stretch their positions on the vertical axis by between 90\% and 110\%. While approaching the problem from a different angle, \texttt{stretch} is computationally similar to \texttt{regress}, except that it emphasizes systematic offset issues rather than systematic slope issues.

@@

DTW is typically useful when you have two sequences, not necessarily of the same length, and you want to (a) calculate how similar they are (as is the case in the \texttt{compare} algorithm) or (b) align the two sequences by mapping each element in one sequence to a corresponding element in the other. For example, DTW may be used to calculate the similarity between a signature, which can be expressed as a sequence of \textit{xy}-coordinates over time, and a reference signature \parencite[e.g.,][]{Lei:2005,Riesen:2018}. Importantly, the two sequences do not need to be perfectly matched in terms of overall magnitude or patterns of acceleration and deceleration for a good alignment to be found. In the case of signature verification, for example, it does not matter if the candidate signature has the same size as the reference or that it was drawn at the same speed, what matters is that there is a good match in the overall shape and that the strokes were drawn in the same order. DTW finds many other applications in, for example, genomics \parencite{Aach:2001}, medicine \parencite{Caiani:1998}, and robotics \parencite{Vakanski:2014}.

@@

If the final fixation on line~$i$ were mapped to the first word on line~$i+1$, this would result in a large increase in the overall cost of the mapping, so line changes act as major clues about the best alignment. The upshot of this is that \texttt{warp} effectively segments the fixation sequence into exactly $m$ subsequences, which are mapped to the lines of text in chronological order. In this sense, \texttt{warp} behaves very much like \texttt{segment}. However, the additional benefit of \texttt{warp} is that it can simultaneously consider different possibilities about which saccades are the return sweeps, selecting only those that result in the best fit to the passage at a global level. Nevertheless, \texttt{warp} is ultimately limited by the veracity of the expected fixation sequence, which encodes one particular way of reading the passage---line by line from start to end. If the reader deviates from this assumption (e.g., by rereading or skipping lines), \texttt{warp} can fail to correctly assign fixations to lines.

@@

For each phenomenon, we ran 100 simulations for each of 50 gradations in the parameter space, and each of these 5000 simulated reading scenarios was corrected by all ten algorithms. Accuracy is measured as the percentage of fixations that were correctly mapped back to the target line. Before describing the results, there are three important things to note. Firstly, the extreme values we have chosen for each phenomenon are arbitrary, so the algorithms should only be compared within, and not across, phenomena.\footnote{For example, \texttt{regress} appears to have worse performance on shift compared to slope; however, if we had simulated a narrower range of shift values, the results might have led us to the opposite conclusion.} Secondly, we have not modeled the interactions between phenomena because it is inherently difficult to explore the effects of five dimensions on accuracy and it is not clear how the dimensions should be weighted a~priori. Thirdly, for the algorithms that have free parameters (\texttt{chain}, \texttt{compare}, \texttt{merge}, \texttt{regress}, and \texttt{stretch}), we use the default parameter settings defined in the previous section. We have not systematically manipulated the parameter settings because (a) this would result in an explosion in the number of algorithm/parameter combinations that we must consider, (b) manipulating a parameter to deal with one phenomenon can have unexpected consequences for other phenomena,\footnote{For example, if the standard deviation bounds of \texttt{regress} are widened, it may be possible to improve performance on noise, but the algorithm will be less capable of dealing with slope.} and (c), in a sense, these algorithms ought to incur a penalty for not being parameter-free.

@@

In terms of shift (Fig.~\ref{fig04}C), when the parameter is zero, all algorithms perform perfectly, but as it becomes more extreme, five of the algorithms---\texttt{attach}, \texttt{chain}, \texttt{regress}, \texttt{split}, and \texttt{stretch}---drop in accuracy. In fact, \texttt{attach}, \texttt{chain}, and \texttt{split} produce identical results in the case of shift because they are all fundamentally reliant on absolute position. Somewhat surprisingly, \texttt{stretch} does not perform especially well on shift. This is because \texttt{stretch} can only handle up~to one full line of shift; any more than this and the bounds have to be relaxed, but this results in an objective function with multiple maxima which is difficult to optimize. The \texttt{compare} algorithm is mostly resilient to shift, while the remaining four algorithms---\texttt{cluster}, \texttt{merge}, \texttt{segment}, and \texttt{warp}---are invariant.

@@

The cleaned sample dataset was corrected independently by two human correctors (JWC and VNP). To perform the correction, each corrector studied plots of the participants' fixation sequences and recorded, fixation by fixation, which line each one belonged to, guided by fixation position, saccade trajectories, textual cues, and fixation duration, as well as general knowledge of eye tracking and reading behavior. Unlike the algorithms, the human correctors also had the option to discard fixations as they saw fit. This is because there were cases where it was clear a fixation should be discarded---for instance, due to spatial misplacement or ultra-short duration---and it would have therefore felt disingenuous to assign these cases to a line anyway.

@@

As with the simulations, accuracy is measured as the percentage of fixations that the algorithm mapped to the correct line; the ground truth is defined by the gold standard manual correction. In cases where the correctors chose to discard a fixation, the algorithm is automatically wrong, which amounts to a constant baseline level of error that all algorithms suffer from equally. Fig.~\ref{fig07} plots accuracy on the 48 sample trials by algorithm. The most striking result is \texttt{compare} with overall median accuracy of 61.2\%, substantially worse than all other algorithms.\footnote{This concurs with the 60\% accuracy reported by \textcite[p.~1231]{LimaSanches:2015}. \textcite[p.~104]{Yamaya:2017}, who describe a slightly more complex method with better sweep detection, report accuracy of 87\%, which is still quite low compared to the other results obtained in this paper.} This contrasts with our simulations, which indicated that \texttt{compare} should at~least be relatively strong on distortion. The reason for this discrepancy is that the simulated fixation sequences were generated directly from the lines of text with one fixation per word, so the artificial gaze lines that \texttt{compare} identified tended to be very horizontally similar to the artificial text lines. In the natural dataset, however, this is not the case; when the data contains a~lot of natural noise and regressions, gaze lines cannot be reliably matched to text lines based on similarity, even if the set of candidate text lines is narrowed down to the three closest neighbors. Given that \texttt{compare} exhibited such poor performance, we consider it to be an algorithmic deadend and do not discuss it any further.

@@

It is also worth considering what proportion of trials are likely to be usable following the application of a particular method. Fig.~\ref{fig10} reports the proportion of trials that obtained accuracy of 90\%, 95\%, and 99\% by algorithm. If you are willing to accept relatively low accuracy at the trial level (e.g., 90\%, Fig~\ref{fig10}A), then \texttt{cluster}, \texttt{merge}, and \texttt{stretch} will provide the best performance---a large proportion of trials will meet this criterion. In comparison, if you have more stringent accuracy requirements at the trial level (e.g., 99\%, Fig~\ref{fig10}C), then \texttt{segment}, \texttt{split}, and \texttt{warp} are likely to provide better performance. Of course, the cost of a more stringent accuracy criterion is that fewer trials will be usable overall, and it is not possible to know which trials have low accuracy in the absence of manual correction data. This highlights the fact that it is currently not possible to confidently achieve a high level of accuracy in a high proportion of trials, so researchers may still need to invest a significant amount of time if a high level of accuracy is demanded.

@@

The analyses presented in this paper clearly indicate that each algorithm performs best on a different set of factors, some of which we have not considered here in detail. For example, if the line spacing is quite tight, the eye tracking data is more likely to be negatively impacted by distortion, making a sequential algorithm a better choice; conversely, if lines are spread far apart, a relative-positional algorithm may be more appropriate. Overall, different sets of data will require different correction algorithms, and a qualitative inspection of the data will be required to detect the relative severity of general noise, drift issues, and regression phenomena. To help in this process, one option might be to hand-correct a sample of the trials in order to assess which algorithm performs best on those specific cases and then apply this algorithm to the entire dataset. However, there might very well be too much trial-by-trial (or participant-by-participant) variability to use a single algorithm across an entire dataset. In this case, it may be preferable to create subsets of data exhibiting comparable patterns of eye tracking phenomena and deal with those subsets with different algorithms. Another idea would be to run several algorithms over the dataset and manually inspect cases where there is disagreement.

@@

Another important aspect to consider is the presence of free parameters. The \texttt{chain}, \texttt{merge}, \texttt{regress}, and \texttt{stretch} algorithms take additional input parameters that must be set appropriately by the user. In practice, \possessivecite{Spakov:2019} and \possessivecite{Cohen:2013} suggested defaults for the \texttt{merge} and \texttt{regress} algorithms seemed to work well on our test dataset, but in the case of \texttt{chain}, it was somewhat unclear how to set the \textit{x}~and \textit{y}~thresholds appropriately, so experimentation might be required to produce the best results. We also found that \texttt{stretch} was very sensitive to its parameter settings and that the upper and lower bounds must be tightly constrained around likely values for it to produce sensible results. An advantage of all other algorithms is that they are parameter free, making drift correction easier to perform, document, and justify.

@@

We would not wish to claim that the algorithms, as presented here, are the only approaches one may take nor that they are the ultimate form of each core method; all can be improved in one way or another. Furthermore, there are likely to be ways of combining the outputs of multiple algorithms to increase confidence in particular solutions. The goal of this paper, however, was to evaluate the algorithms in their more abstract, idealized forms in order to make general recommendations and to provide a solid foundation for the future development of vertical drift correction software. Nevertheless, here we briefly note some of the most obvious ways in which the algorithms could be improved.

@@

The biggest weakness of \texttt{cluster} was its ability to deal with general noise (or, equivalently, tight line spacing). One potential way to improve this would be to utilize the \textit{x}~values of the fixations. Unfortunately, it is not simply a case of performing a two-dimensional \textit{k}-means clustering on the \textit{xy}~values because this leads to situations where clusters are identified that span multiple lines because they have similar \textit{x}~values. However, it might still be possible to utilize the \textit{x}-axis information, perhaps by weighting the two axes differently in some way. Cluster analysis is a very broad topic in data science, and there are likely to be many other candidate algorithms, beyond simple \textit{k}-means clustering, that will be worth investigating.

@@

Our analyses showed that \texttt{stretch} behaves very similarly to \texttt{regress}. This is because they are essentially two variants on the same basic idea: Detect the magnitude of the underlying slope (\texttt{regress}) or shift (\texttt{stretch}) and then reverse it. However, this does not work so well if the underlying calibration error is fluctuating in time or space. One way to improve the method, then, would be to search a more complex transformation space by including rotations and shears, for example, or by applying separate transformations to each quadrant of the text. Additionally, as Equation~\ref{equation1} makes clear, \texttt{stretch} essentially has the \texttt{attach} algorithm embedded within it, but in principle it should be possible to substitute this with any of the positional algorithms. For example, a \texttt{stretch-chain} algorithm would find a transformation of the fixations that results in minimal change when you apply the \texttt{chain} algorithm.
